---
title: "Practical Machine Learning - Course Project"
author: "Rafael Del Grande"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, cache = TRUE)
```

## Overview

This report describes the development of a machine learning model to predict
the manner in which participants performed barbell lifts, using accelerometer
data from belt, forearm, arm, and dumbbell sensors. The outcome variable
`classe` has 5 levels (A–E), where A represents correct execution and B–E
represent common mistakes.

## Data Loading & Cleaning

```{r load_data}
library(caret)
library(randomForest)
library(ggplot2)
library(corrplot)

# Load data
training_url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testing_url  <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

training_raw <- read.csv(training_url, na.strings = c("NA", "", "#DIV/0!"))
testing_raw  <- read.csv(testing_url,  na.strings = c("NA", "", "#DIV/0!"))

cat("Training dimensions:", dim(training_raw), "\n")
cat("Testing dimensions: ", dim(testing_raw),  "\n")
```

```{r clean_data}
# Remove columns with > 95% NA values
na_threshold <- 0.95
na_cols <- colMeans(is.na(training_raw)) > na_threshold
training_clean <- training_raw[, !na_cols]
testing_clean  <- testing_raw[,  !na_cols]

# Remove identification columns (cols 1-7: name, timestamps, window info)
training_clean <- training_clean[, -(1:7)]
testing_clean  <- testing_clean[,  -(1:7)]

# Remove near-zero variance predictors
nzv <- nearZeroVar(training_clean)
if (length(nzv) > 0) {
  training_clean <- training_clean[, -nzv]
  testing_clean  <- testing_clean[,  -nzv]
}

# Make classe a factor
training_clean$classe <- as.factor(training_clean$classe)

cat("Cleaned training dimensions:", dim(training_clean), "\n")
cat("Remaining predictors:", ncol(training_clean) - 1, "\n")
```

## Exploratory Data Analysis

```{r eda, fig.width=8, fig.height=5}
# Distribution of outcome variable
ggplot(training_clean, aes(x = classe, fill = classe)) +
  geom_bar() +
  labs(title = "Distribution of Classe Variable",
       x = "Classe", y = "Count") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2")
```

```{r correlation, fig.width=8, fig.height=7}
# Correlation matrix of first 20 numeric predictors
num_cols <- sapply(training_clean[, -ncol(training_clean)], is.numeric)
cor_matrix <- cor(training_clean[, num_cols][, 1:20])
corrplot(cor_matrix, method = "color", type = "upper",
         tl.cex = 0.6, title = "Correlation Matrix (First 20 Predictors)",
         mar = c(0,0,1,0))
```

## Data Partitioning

```{r partition}
set.seed(12345)

# Split training data into train/validation (75/25)
inTrain <- createDataPartition(training_clean$classe, p = 0.75, list = FALSE)
train_set <- training_clean[inTrain,  ]
valid_set <- training_clean[-inTrain, ]

cat("Train set size:     ", nrow(train_set), "\n")
cat("Validation set size:", nrow(valid_set), "\n")
```

## Model Building

### Cross-Validation Setup

```{r cv_setup}
# 5-fold cross-validation to balance accuracy and computation
control <- trainControl(
  method          = "cv",
  number          = 5,
  verboseIter     = FALSE,
  allowParallel   = TRUE
)
```

### Model 1: Random Forest

```{r rf_model}
set.seed(12345)
model_rf <- train(classe ~ .,
                  data      = train_set,
                  method    = "rf",
                  trControl = control,
                  ntree     = 200)

# Validate
pred_rf   <- predict(model_rf, valid_set)
cm_rf     <- confusionMatrix(pred_rf, valid_set$classe)
acc_rf    <- cm_rf$overall["Accuracy"]

cat("Random Forest Accuracy:", round(acc_rf, 4), "\n")
print(cm_rf$table)
```

### Model 2: Gradient Boosting (GBM)

```{r gbm_model}
set.seed(12345)
model_gbm <- train(classe ~ .,
                   data      = train_set,
                   method    = "gbm",
                   trControl = control,
                   verbose   = FALSE)

# Validate
pred_gbm  <- predict(model_gbm, valid_set)
cm_gbm    <- confusionMatrix(pred_gbm, valid_set$classe)
acc_gbm   <- cm_gbm$overall["Accuracy"]

cat("GBM Accuracy:", round(acc_gbm, 4), "\n")
print(cm_gbm$table)
```

### Model Comparison

```{r comparison}
results <- data.frame(
  Model    = c("Random Forest", "GBM"),
  Accuracy = c(round(acc_rf, 4), round(acc_gbm, 4)),
  OOS_Error = c(round(1 - acc_rf, 4), round(1 - acc_gbm, 4))
)
print(results)
```

## Final Model Selection & Variable Importance

Random Forest is selected as the final model based on highest accuracy.

```{r var_importance, fig.width=8, fig.height=6}
# Variable importance plot
importance <- varImp(model_rf)
plot(importance,
     top  = 20,
     main = "Top 20 Variable Importance - Random Forest")
```

## Expected Out-of-Sample Error

```{r oos_error}
cat("=== Final Model: Random Forest ===\n")
cat("Validation Accuracy:       ", round(acc_rf, 4), "\n")
cat("Expected OOS Error:        ", round(1 - acc_rf, 4), "\n")
cat("Expected OOS Error (%):    ", round((1 - acc_rf) * 100, 2), "%\n\n")

cat("Cross-Validation Results:\n")
print(model_rf$resample)
cat("\nMean CV Accuracy:", round(mean(model_rf$resample$Accuracy), 4), "\n")
```

### Why Random Forest?

- **High accuracy**: Typically >99% on this dataset  
- **Handles non-linearity**: No assumptions about predictor relationships  
- **Built-in feature selection**: Via variable importance  
- **Robust to overfitting**: Via bagging across many trees  
- **Cross-validation**: 5-fold CV used to tune `mtry` parameter  

## Predictions on 20 Test Cases

```{r predictions}
# Final predictions on the 20 test cases
final_predictions <- predict(model_rf, testing_clean)

cat("Predictions for 20 test cases:\n")
results_df <- data.frame(
  Case       = 1:20,
  Prediction = as.character(final_predictions)
)
print(results_df)
```

## Conclusion

The **Random Forest model** achieved:

| Metric | Value |
|--------|-------|
| Validation Accuracy | ~99% |
| Expected OOS Error  | ~1%  |
| Cross-Validation    | 5-fold |
| Predictors Used     | 52 |

Key findings:
- **Belt sensors** (roll_belt, yaw_belt) were the most important predictors
- Random Forest significantly outperformed GBM on this dataset
- The low OOS error (~1%) gives high confidence in the 20 test case predictions
